---
title: "Getting Started: Initializing an AquaCrop Project"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting Started: Initializing an AquaCrop Project}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Quick Start

The setup workflow is three steps: load, initialize, reload.

**1. Load the package**

```{r library}
library(aquacropr)
```

**2. Initialize the project**

`init_aquacrop()` creates the full directory structure and downloads the
AquaCrop binary in one call. By default (`version = NULL`), it always
fetches the latest officially released binary for your platform — no
compiler required:

```{r init}
init_aquacrop(path = "~/my_project")
```

**3. Reload after RStudio switches sessions**

Once initialization is complete, RStudio automatically opens a fresh
session inside the new project directory. Simply reload the package to
continue your work:

```{r reload}
library(aquacropr)
```

You are now ready to write input files and run simulations with
`run_aquacrop()`.

## Choosing an AquaCrop Version

The `version` argument gives you full control over which binary is
installed.

### Latest release `(default)`

`version = NULL` always downloads the most recent stable binary
published by FAO. This is the recommended option for most users: it
requires no compilation and guarantees access to the latest AquaCrop
features the moment they are officially released.

```{r version-null}
init_aquacrop(path = "~/my_project")
```

### Pinned release

Pass an explicit version string when reproducibility matters — for
instance, to match a binary used in a published study:

```{r version-pin}
init_aquacrop(path = "~/my_project", version = "7.1")
```

Note : When `init_aquacrop()` downloads a binary, it saves it to a local cache directory on your machine. The next time you call `init_aquacrop()` with the same version, the binary is loaded directly from cache — no internet connection needed, no waiting. The cache is keyed by version string, so `"7.1"` and `"7.2"` are stored independently. Switching versions is instant if you have already downloaded them before. The only way to bypass the cache and force a fresh download is to pass `force = TRUE`.

### Development version

`version = "dev"` compiles AquaCrop directly from the latest `main`
branch source code. Use this when you need features that have not yet
made it into an official release. A Fortran compiler is required (see
[Installing from Source](#installing-aquacrop-from-source) below):

```{r version-dev}
init_aquacrop(path = "~/my_project", version = "dev")
```


## Additional Options

### Overwriting an existing project

By default, `init_aquacrop()` will not touch an existing installation.
Use `force` to reinstall the binary and `overwrite` to reset the
directory structure:

```{r force}
init_aquacrop(path = "~/my_project", force = TRUE, overwrite = TRUE)
```

### Skipping the RStudio project file

An `.Rproj` file is created at the project root by default. Set
`use_rproject = FALSE` to opt out:

```{r rproject}
init_aquacrop(path = "~/my_project", use_rproject = FALSE)
```

### Cross-platform projects

The operating system is auto-detected at runtime. You can override this
with the `os` argument — useful when preparing a project on one machine
that will run on another:

```{r os}
# Build a Windows-ready project from Linux or macOS
init_aquacrop(path = "~/my_project", os = "windows")
```


## Installing AquaCrop from Source {#installing-aquacrop-from-source}

When `version = "dev"` is used, `init_aquacrop()` internally delegates
to `install_source()`, which runs the following steps automatically:

1.  **Compiler detection** — queries `R CMD config FC` once and reuses the result throughout. Override with the `compiler` argument if needed.
2.  **Source download** — fetches the latest `main` branch archive from the AquaCrop repository.
3.  **Compilation** — invokes `make` with the detected Fortran compiler.
4.  **Installation** — copies the compiled executable to `path`.
5.  **Cleanup** — removes temporary build files unless
    `keep_source = TRUE`.

System requirements by platform:

| Platform | Required tools |
|--------------------------------|----------------------------------------|
| **Linux** | GNU Make ≥ 3.82 · gfortran ≥ 6.4.0 *or* ifort ≥ 18.0.1 |
| **macOS** | GNU Make ≥ 3.82 · gfortran from [mac.r-project.org/tools](https://mac.r-project.org/tools) |
| **Windows** | [Rtools](https://cran.r-project.org/bin/windows/Rtools/) |

> **macOS note:** The `clang` compiler bundled with Xcode does not
> include Fortran support. Always use the gfortran binary from the
> official R toolchain at mac.r-project.org/tools.

## Project Directory Structure

A successful call to `init_aquacrop()` produces the following layout
under `path`:

```         
my_project/
├── my_project.Rproj      # RStudio project file (if use_rproject = TRUE)
├── aquacrop[.exe]        # AquaCrop executable
├── README.txt            # Project notes
├── CAL/                  # Calendar files (.CAL)
├── CLIMATE/              # Climate input files (.Tnx, .ETo, .PLU, .CLI, .CO2)
├── CROP/                 # Crop parameter files (.CRO)
├── LIST/                 # Project Master files (.PRM)
├── MANAGEMENT/           # Field management files: (.MAN, .IRR)
├── OBS/                  # Observed field data for simulation evaluation
├── OUTP/                 # Simulation output files
├── PARAM/                # AquaCrop program parameter files
├── SIMUL/                # Simulation settings
└── SOIL/                 # Soil profile files (.SOL, .SW0, .GWT)
```

## Relationship to the Official AquaCrop Standalone

Users already familiar with the [FAO AquaCrop standalone
(plug-in)](https://www.fao.org/aquacrop/en) will recognize the structure
immediately. `aquacropr` preserves all four core directories of the official standalone (`LIST`, `PARAM`, `SIMUL`, `OUTP`) and extends the design by replacing the historical catch-all `DATA/` folder with clearly typed, dedicated directories — one per input category.

| Directory     | AquaCrop Standalone v7.1      | `aquacropr`    |
|---------------|-------------------------------|--------------------|
| `LIST/`       | ✅ Project files (.PRO, .PRM) | ✅                 |
| `SIMUL/`      | ✅ Simulation settings        | ✅                 |
| `PARAM/`      | ✅ Program parameters         | ✅                 |
| `OUTP/`       | ✅ Output files               | ✅                 |
| `OBS/`        | ✅ Observed field data        | ✅                 |
| `CLIMATE/`    | ❌ (merged in `DATA/`)        | ✅ Explicit folder |
| `CROP/`       | ❌ (merged in `DATA/`)        | ✅ Explicit folder |
| `SOIL/`       | ❌ (merged in `DATA/`)        | ✅ Explicit folder |
| `MANAGEMENT/` | ❌ (merged in `DATA/`)        | ✅ Explicit folder |
| `CAL/`        | ❌ (merged in `DATA/`)        | ✅ Explicit folder |

This structure makes large-scale simulation projects easier to navigate,
script, and version-control, while remaining fully compatible with the
input file formats expected by the AquaCrop executable.
