---
title: "Onset of the Growing Season: Methods, Limitations, and Sahelian Realities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Onset of the Growing Season: Methods, Limitations, and Sahelian Realities}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Background and design rationale

### A limitation of the AquaCrop standalone

**On the `.CAL` file.** 
The `.CAL` file works as intended in the
AquaCrop GUI, which appears to use it to derive sowing dates before
writing the project master file (`.PRM`). In standalone mode, the
exact mechanism is not documented — but the sowing date embedded in
the `.PRM` is the only input AquaCrop needs to anchor the growing
period. Whether or not the standalone re-evaluates the onset rule at
runtime is therefore, in practice, irrelevant.



### The `aquacroptools` approach

What `aquacroptools` does is straightforward: `find_onset()` reads the
`.CAL`, computes the sowing date series from the climate data, and
`write_prm()` writes those dates directly into the `.PRM`. The
standalone receives exactly what it expects. By owning this
computation in R, `aquacroptools` is free to implement any onset rule
that can be expressed as a function of daily climate data — the four
native AquaCrop criteria are fully reproduced, and three additional
methods designed for rainfall-limited environments are added on top.

**You do not need to call `find_onset()` for simulations to work.**
When you pass a `.CAL` to `write_prm()` and launch a simulation with
`run_aquacrop()`, onset dates are computed and embedded automatically.
`find_onset()` is provided so you can inspect, validate, and compare
onset dates before committing to a configuration — a transparent window
into the planting-date engine, not a required step.

The typical workflow is:

```{r workflow}
library(aquacroptools)

# 1. Choose and encode your onset rule
write_cal(
  site_name       = "Ouaga",
  onset           = "rainfall",
  window_start    = 100,
  window_length   = 120,
  criterion       = 2,
  preset_value    = 20,
  successive_days = 3
)

# 2. (Optional but recommended) Inspect the resulting onset dates
find_onset("Ouaga")
#>   year onset_doy onset_date
#> 1 1981       147 1981-05-27
#> 2 1982       153 1982-06-02
#> ...

# 3. Pass the .CAL to write_prm() — everything else is automatic
write_prm(site_name = "Ouaga", ...)
```



## Novel implementations in `aquacroptools`

The three Sahelian methods below are **exclusively available in
`aquacroptools`**. They are not accessible through the AquaCrop
standalone GUI and are not part of the official AquaCrop criterion
set.

| Method | Origin | Available in AquaCrop standalone |
|---|---|---|
| Native criteria 1–4 (rainfall) | AquaCrop / FAO | ✅ Yes |
| Native criteria 1–4 (thermal) | AquaCrop / FAO | ✅ Yes |
| **Sivakumar (1988)** | Agroclimatology literature | ❌ `aquacroptools` only |
| **Marteau et al. (2009)** | Agroclimatology literature | ❌ `aquacroptools` only |
| **Fuzzy logic — Criterion 5 (Waongo et al., 2014)** | Agroclimatology literature | ❌ `aquacroptools` only |

Their inclusion is a direct consequence of the reimplementation
strategy: once onset detection is in R, adding a new method is a matter
of implementing a new decision function — no recompilation, no GUI
patch required.


## Why onset matters

In rainfed agriculture, the onset of the growing season is the most
consequential decision a farmer makes each year. Sow too early and a
dry spell after germination can be fatal. Sow too late and the season
is too short to reach maturity. In the Sahel, where the rainy season
lasts 90 to 150 days and rainfall is highly variable in time and space,
the planting date alone can explain yield differences of 20% or more
between identical fields in the same year (Waongo et al., 2014).

Getting the onset rule right is therefore not a technical detail — it
is the agronomic foundation of your simulation.


## The three onset types

`write_cal()` supports three `onset` values, each determining which
climate file `find_onset()` will read at runtime.

| `onset`      | Climate file        | Typical use case                          |
|--------------|---------------------|-------------------------------------------|
| `"fixed"`    | `.PLU` (years only) | Fixed sowing calendar, irrigated systems  |
| `"rainfall"` | `.PLU`              | Rainfed crops, semi-arid to sub-humid     |
| `"thermal"`  | `.Tnx`              | Temperate crops, frost- or heat-triggered |

For sub-Sahelian contexts, `onset = "rainfall"` is the appropriate
choice in virtually all situations.



## Native AquaCrop rainfall criteria (1–4)

### Criterion 1 — Cumulative rainfall since the start of the window

Onset is triggered as soon as cumulative rainfall since `window_start`
reaches `preset_value` mm.

```{r crit1}
write_cal(
  site_name     = "Ouaga",
  onset         = "rainfall",
  window_start  = 100,
  window_length = 120,
  criterion     = 1,
  preset_value  = 25
)
```

A single heavy convective event can trigger onset even if the following
two weeks are completely dry — a common pattern in the northern Sahel.
Best suited to wetter agroclimatic zones where the onset, once
triggered, tends to be sustained.



### Criterion 2 — Observed rainfall over N consecutive days

Onset is triggered when total rainfall over `successive_days`
consecutive days equals or exceeds `preset_value` mm.

```{r crit2}
write_cal(
  site_name       = "Ouaga",
  onset           = "rainfall",
  window_start    = 100,
  window_length   = 120,
  criterion       = 2,
  preset_value    = 20,
  successive_days = 3
)
```

This mirrors the agronomic rule used by extension services in Burkina
Faso and Mali: ≥ 20 mm over 3 days provides sufficient soil moisture
for germination of maize and sorghum. Setting `occurrences = 2`
requires the condition to be met twice before onset is declared — a
simple but effective filter against isolated convective events.



### Criterion 3 — 10-day (dekadal) rainfall

Onset is triggered when accumulated rainfall over the 10-day dekad
containing day *j* equals or exceeds `preset_value` mm. Operationally
aligned with the dekadal bulletins produced by national meteorological
services across West Africa, including ANAM-BF.

```{r crit3}
write_cal(
  site_name     = "Ouaga",
  onset         = "rainfall",
  window_start  = 100,
  window_length = 120,
  criterion     = 3,
  preset_value  = 30
)
```



### Criterion 4 — 10-day rainfall as a fraction of 10-day ETo

Requires an `.ETo` file. Onset is triggered when 10-day rainfall
exceeds `preset_value` percent of the 10-day reference
evapotranspiration. Accounts for atmospheric demand — conceptually
superior to rainfall-only thresholds in semi-arid environments, but
requiring careful calibration given that pre-monsoon ETo in the Sahel
can reach 8–10 mm/day.

```{r crit4}
write_cal(
  site_name     = "Ouaga",
  onset         = "rainfall",
  window_start  = 100,
  window_length = 120,
  criterion     = 4,
  preset_value  = 50
)
```



## Novel Sahelian criteria

All three methods below address a common gap in the native AquaCrop
criteria: none of the built-in rules were designed with the specific
rainfall dynamics of the Sahel in mind — namely, the frequency of false
starts, the importance of post-germination dry spell risk, and the need
for methods validated against farmer behaviour in the field.

### Sivakumar (1988)

Sivakumar (1988) proposed one of the earliest operationally robust
definitions of onset for West Africa: the first day after a threshold
date when rainfall over a given number of days meets a minimum,
provided no prolonged dry spell follows within a defined subsequent
period. Built from long-term station data across the Sudanian and
Sahelian zones, this definition directly captures the agronomic concern
that a good start is one that holds.


### Marteau et al. (2009)

Marteau et al. (2009) refined the Sivakumar definition using field
observations of actual sowing decisions by pearl millet farmers in
southwest Niger. Their formulation adjusts rainfall and dry spell
thresholds based on agronomic field evidence rather than meteorological
convention alone, anchoring the onset definition in observed farmer
behaviour. It is particularly well suited for operational use in the
Sahelian belt.



### Criterion 5 — Fuzzy logic (Waongo et al., 2014)

Criteria 1 through 4, as well as the Sivakumar and Marteau approaches,
share a structural limitation: they are **binary**. A cumulative
rainfall of 24.9 mm does not trigger a criterion set at 25 mm, even
though agronomically the two situations are nearly identical. This
artificial cliff effect tends to exaggerate interannual variability in
simulated planting dates and is particularly problematic in
probabilistic forecasting contexts.

Criterion 5 addresses this by replacing hard thresholds with **three
fuzzy membership functions**, each mapping a physical indicator to a
planting suitability score in $[0, 1]$:

| Component | Symbol | What it measures | Direction |
|---|---|---|---|
| Cumulative rainfall over `accum_days` days | $\gamma_1$ | Soil moisture recharge | Increasing ↑ |
| Number of wet days over `accum_days` days | $\gamma_2$ | Temporal distribution of rain | Increasing ↑ |
| Longest dry spell in the 30 days after onset | $\gamma_3$ | Post-germination risk | **Decreasing ↓** |

Each function is trapezoidal, defined by a lower and upper bound.
Onset is declared on the first day $j$ for which:

$$\gamma_1(j) \times \gamma_2(j) \times \gamma_3(j) \geq \text{threshold}$$

Because the index is a product, all three components must contribute
simultaneously. Abundant rain concentrated in a single event
($\gamma_2$ low), or rain immediately followed by a long dry spell
($\gamma_3$ low), will not trigger onset even if the cumulative total
is high. This makes criterion 5 the only method in `aquacroptools`
that explicitly and continuously penalises post-planting dry spell
risk.

```{r crit5}
write_cal(
  site_name       = "Ouaga",
  onset           = "rainfall",
  window_start    = 100,
  window_length   = 120,
  criterion       = 5,
  preset_value    = 20,    # cum_rain_lower (mm)
  successive_days = 3,     # accum_days
  cum_rain_upper  = 40,
  wet_days_lower  = 1,
  wet_days_upper  = 3,
  dry_spell_lower = 7,
  dry_spell_upper = 15,
  fuzzy_threshold = 0.5
)
```

All eight parameters are written into the `.CAL` file and recovered
automatically when `find_onset()` or `write_prm()` is called.



## Thermal criteria

Native to AquaCrop, thermal criteria use the `.Tnx` file and are
designed for temperature-limited growing seasons. They are not relevant
for the Sahel but are fully available when `aquacroptools` is used in
temperate or highland contexts.

| Criterion | Trigger condition |
|-----------|-------------------|
| 1 | Daily Tmin ≥ preset for N successive days |
| 2 | Daily Tmean ≥ preset for N successive days |
| 3 | Cumulative GDD ≥ preset over N successive days |
| 4 | Cumulative GDD since window start ≥ preset |



## Fixed onset

When the sowing date is prescribed — irrigated schemes, experimental
designs, or climatological baselines — `onset = "fixed"` is the right
choice.

```{r fixed}
write_cal(
  site_name = "Bama_irrigated",
  onset     = "fixed",
  fixed_day = 166   # June 15
)
```

> `find_onset()` reads `CLIMATE/<site_name>.PLU` from the working
> directory. For multi-method comparisons on the same station, copy or
> symlink the same `.PLU` under each `site_name`.



## Sahelian realities

### False starts

The Sahelian rainy season is characterised by high intra- and
inter-annual variability. False starts — isolated rainfall events
followed by dry spells of 10 days or more — are frequent in May and
early June, particularly north of the 12th parallel (Sultan et al.,
2005). Criteria 1 and 3 do not penalise post-onset dry spells and are
the most exposed to this risk. Criterion 2 with `occurrences = 2`, or
any of the three Sahelian-specific methods, provide progressively
stronger protection.

### Indicative parameter ranges for criterion 5

Starting points derived from Waongo et al. (2014) and operational
practice at ANAM-BF. Local calibration against observed planting dates
is always recommended before operational use.

| Parameter | Sudan-Sahel (700–900 mm/yr) | Sahel (400–700 mm/yr) |
|---|---|---|
| `preset_value` (cum_rain_lower) | 15–20 mm | 10–15 mm |
| `cum_rain_upper` | 30–40 mm | 25–35 mm |
| `accum_days` | 3 days | 3 days |
| `wet_days_lower` | 1 | 1 |
| `wet_days_upper` | 3 | 2–3 |
| `dry_spell_lower` | 7–10 days | 5–7 days |
| `dry_spell_upper` | 15–20 days | 10–15 days |
| `fuzzy_threshold` | 0.4–0.6 | 0.3–0.5 |
| `window_start` | DOY 100 (Apr 10) | DOY 120 (Apr 30) |
| `window_length` | 120 days | 100 days |

### Fallback behaviour

When no day in the search window satisfies the onset criterion,
`find_onset()` returns the last day of the window
(`window_start + window_length - 1`). This sentinel value ensures that
simulations always receive a valid planting date. Years where the
fallback is triggered should be flagged in your analyses as "no onset
detected" and not interpreted as genuine late onsets.



## References

Marteau, R., Sultan, B., Moron, V., Guilderson, T., Kane, C., &
Traoré, S. B. (2009). The onset of the rainy season and farmers'
sowing strategy for pearl millet cultivation in Southwest Niger.
*Agricultural and Forest Meteorology*, 148(10), 1738–1750.

Sivakumar, M. V. K. (1988). Predicting rainy season potential from the
onset of rains in Southern Sahelian and Sudanian climatic zones of
West Africa. *Agricultural and Forest Meteorology*, 42(4), 295–305.

Sultan, B., Baron, C., Dingkuhn, M., Sarr, B., & Janicot, S. (2005).
Agricultural impacts of large-scale variability of the West African
monsoon. *Agricultural and Forest Meteorology*, 128(1–2), 93–110.

Waongo, M., Laux, P., Traoré, S. B., Sanon, M., & Kunstmann, H.
(2014). A crop model and fuzzy rule based approach for optimizing maize
planting dates in Burkina Faso, West Africa. *Journal of Applied
Meteorology and Climatology*, 53(3), 598–613.
